Tulai

### 2023-09-18 paper/paper_tulai.qmd
edited and polished the codes. Good to render now.

```{r}
# Generate a summary table by grouping by 'depth'
summary_table_depth_tool_type <- tl_final %>%
  group_by(depth) %>%
  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE))

# Create a full summary table by joining it back to your original summary table based on 'depth'
summary_table_depth_area_layer <- 
  summary_table_depth_area_layer %>%
  left_join(summary_table_depth_tool_type, 
            by = "depth")

# Print column names 
print(names(summary_table_depth_area_layer))

@@

google_sheet_url <- 
  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"

@@

```{r}
# Create the plot: WABI_Level_logscale
# here I do not consider those levels with a value 0 of retouch frequency. 
WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  geom_text(aes(label = label), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

@@

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, I consider the thickness; is that OK?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
##core frequency
##blank frequency
##retouch frequency
##tool diversity
##thickness of deposit
##lithic density

@@

```{r}
pca_data <- sum_depth_sheet %>% 
  select(thickness, lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
  na.omit()  # Remove rows with NA values

@@

```{r}
library(FactoMineR)
library(factoextra)
# Assuming 'sum_depth_sheet' is your dataframe and you have added 'm_index' as your Menhinick index
pca_data <- sum_depth_sheet[, c('thickness', 'lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]

@@

@@

eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage of variance", "cumulative percentage of variance")]

@@

```{r}
# Visualize biplot:1
fviz_pca_biplot(res.pca,
                axes = c(1, 2))
rownames(res.pca$ind$coord) <- sum_depth_sheet$label
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

@@

```{r}
# Visualize biplot: 2
fviz_pca_biplot(res.pca,
                axes = c(3, 4))
rownames(res.pca$ind$coord) <- sum_depth_sheet$label
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

print(fviz_pca_biplot(res.pca, axes = c(3, 4)))

@@

-----------------------------------------------
DONE UPTO HERE
-----------------------------------------------

PCA_al:

```{r}
#PCA: area_layer (al)
#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level. we have 5 stratigraphical layers: e, d, c, b, a
#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
#e: excavated area is A/ depth of excavation is 0-75(mean)
#d: excavated area is A/ 0.75m-3m→ 2.25
#c: d1 and TP → 30 cm and 40cm
#b: TP/ 170 cm 
#a: D1: 70cm

# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
library(dplyr)

summary_table_al_pca <- summary_table_al %>%
  mutate(
    length = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 4,
      al == "TP1_c" ~ 2,
      al == "TA_e" ~ 4,
      TRUE ~ NA_real_
    ),
    width = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 5,
      al == "TP1_c" ~ 1.8,
       al == "TA_e" ~ 5,
      TRUE ~ NA_real_
    ),
    thickness = case_when(
      al == "D1_a" ~ 0.7,
      al == "D1_c" ~ 0.3,
      al == "TA_d" ~ 2.25,
      al == "TP1_c" ~ 0.4,
      al == "TA_e" ~ 0.75,
      TRUE ~ NA_real_)
  ) %>%

  mutate(volume = length * width * thickness) %>%
  relocate(length, width, thickness, volume, .after = al)
# Show the updated data frame
summary_table_al_pca

```{r}
# Load required library
library(dplyr)

# Add a new column called 'blank_count'
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())

# Calculate core frequency, blank frequency, retouch frequency, lithic density 
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(
    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
    core_freq = round(core_count / volume, 2),
    blank_freq = round(blank_count / volume, 2),
    retouch_freq = round(retouched_tool_count / volume, 2)
  )

# Calculate tool diversity using Menhinick's index
summary_table_al_with_diversity <- tl_final %>% 
  filter(retouch == 1) %>%
  group_by(al, `tool-typology`) %>%  
  summarise(n = n(), .groups = 'drop') %>% 
  group_by(al) %>%  
  summarise(
    num_unique_tool_types = n_distinct(`tool-typology`),
    total_retouched_tools = sum(n),
    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
    .groups = 'drop'
  )

# Merge this new information into the existing summary_table_al_pca
summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")

# Remove the 'Total' row and select only the columns needed for PCA
fin_summary_table_al_pca <- summary_table_al_pca %>%
  filter(al != "Total") %>%
  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)

# Show the final table
print(fin_summary_table_al_pca)

```{r}
library(FactoMineR)
# Perform PCA
res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column

```{r}
## Got error: couldnt figure out the reason of error

# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
eigen_df <- data.frame(
  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
  Eigenvalue = round(eigenvalues[, 1], 3),
  `Variance percent` = round(eigenvalues[, 2], 3),
  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
)

# Print the table
print(eigen_df)

```{r}
# variables to PC1
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 1,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 2,  
             top = 10)

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 3,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 4,  
             top = 10)
```

```{r}
# Load the required package
library(kableExtra)

# Create a table with kable and kableExtra
eigen_df %>%
  kable("html") %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(1, bold = T)
```

# Rounding the numerical columns to 3 decimal places
eigen_df[, -1] <- round(eigen_df[, -1], 3)

# Replacing 'comp' with 'Dim'
eigen_df$Dimension <- gsub("comp", "Dim", eigen_df$Dimension)

# Creating the kable output
kable_output <- kable(eigen_df, 
                      "html", 
                      align = 'c', 
                      col.names = c("Dimension", 
                                    "Eigenvalue", 
                                    "Variance percent", 
                                    "Cumulative variance percent")) %>%
  kable_styling("striped", 
                full_width = F) %>%
  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))

htmltools::browsable(kable_output)

```{r}
# Visualize biplot:1
fviz_pca_biplot(res.pca,
                axes = c(1, 2))
rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

```{r}
# Visualize biplot: 2
fviz_pca_biplot(res.pca,
                axes = c(3, 4))
rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

print(fviz_pca_biplot(res.pca, axes = c(3, 4)))

```{r}
# BM: My preference would be to show the calculation of these 
  # values from the raw data, rather than type them in 
  # here, where we can't see where they came from, we can't easily
  # verify these values as they are currently 
##SS: fixed

# Define the volumes for each area

# add the dimensions for each area
dimensions <- data.frame(
  area = c("ta", "tp1", "d1"),
  length = c(4, 2, 1.5),
  width = c(5, 1.8, 1.5),
  thickness = c(3, 1.6, 0.95)
)

# Calculate volume for each area based on the dimensions
dimensions <- dimensions %>%
  mutate(volume = length * width * thickness)

# Count the number of artifacts by area
artifact_counts <- tl_final %>%
  filter(area %in% c("ta", "tp1", "d1")) %>%
  group_by(area) %>%
  tally()

# Join artifact_counts with dimensions (which now includes volume) by area
lithic_density_df <- left_join(artifact_counts, dimensions, by = "area")

# Compute lithic density for each area and round it to 2 decimal places
lithic_density_df <- lithic_density_df %>%
  mutate(lithic_density = round(n / volume, 2)) # number of artifacts / volume

# Create the summary table
summary_table2_area <- tl_final %>%
  filter(area %in% c("ta", "tp1", "d1")) %>%
  group_by(area) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics), 2)
  )

# Merge the density data with the summary table
summary_table2_area <- 
  left_join(summary_table2_area, 
            lithic_density_df[, c("area", "lithic_density")], 
            by = "area") %>%
  mutate(area = recode(area, 'ta' = 'TA', 
                       'tp1' = 'TP1', 
                       'd1' = 'D1'),
         lithic_volumetric_density = lithic_density)

# Remove the 'lithic_density' column
summary_table2_area <- summary_table2_area %>% 
  select(-lithic_density)

# Print out the resulting data frame
print(summary_table2_area)

```{r}
# generate table 2 - area

library(kableExtra)

kable(summary_table2_area, 
      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
      align = 'c',
      col.names = c("Area", 
                    "Number of Retouched Lithics", 
                    "Total Number of Lithics", 
                    "Retouch frequency", 
                    "Lithic Volumetric Density(pcs/(m^3)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
  column_spec(1, width = "1cm") %>% 
  column_spec(2, width = "1cm") %>%

  column_spec(3, width = "1cm") %>% 
  column_spec(4, width = "1cm") %>% 
  column_spec(5, width = "1cm")

------------------------------
Done with the main plots
------------------------------

```{r}
# WABI_LEVEL

# BM: I am confused why we are getting the counts of artefacts 
# and retouch pieces from the google sheet, after you wrote so many lines of
# code above to prepare the data from the CSV file. Where did the lithic data
# in the google sheet come from? My preference is to have 
# only copy of the lithic data in one place, and then join it with 
# other data using code. Currently it seems like we have the 
# lithic data in both the google sheet and the CSV file here. I'm worried that 
# will be confusing and be difficult to update and keep correct. I prefer that we 
# omit the lithic data from the google sheet, then import the google sheet and 
# join it to the lithic data from the CSV file. Then we can have a data frame
# with the area/layer/level data and lithic data to work on for plotting.

#SS: the googlesheet is the last thing I created. I created the summary table in R, and then for splitting up the depth (we have few depth range that was not possible to assign the layer to them, so, we splited them), I found it is easier to do so in Google sheet.

filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)

WABI_Level <- ggplot(filtered_data, 
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 2) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              size = 0.5) +
  geom_text(aes(label = area_layer_level), 
            vjust = 1.5, 
            hjust = 0.5, 
            size = 2,
            check_overlap = TRUE) +
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                 stat(rr.label), 
                                 stat(p.value.label), 
                                 sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +
  # Add titles and labels
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +

  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

WABI_Level

```{r}
# whole assemblage

  # Summarize data by Area
summary_table_tech <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  group_by(area) %>% 
  summarise(
   blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),

    core_fragment_count = sum(`corefrag` == "core fragment", na.rm = TRUE),
    core_rejuvention_count = sum(`corefrag` == "rejuvenation piece", na.rm = TRUE),
    core_tablet_count = sum(`corefrag` == "core tablet", na.rm = TRUE),
    core_crested_count = sum(`corefrag` == "crested bladelet", na.rm = TRUE),
    core_primarypiece_count = sum(`corefrag`  %in% c("primary flake", "primary blade"), na.rm = TRUE),

   total = n()
  )
```

```{r}
# plot_blank_type

# Data filtering and plotting
tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>% 
  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
  ggplot(aes(x = area, fill = blank)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Area",
    y = "Number",
    fill = "Blank Type"
  ) +
  theme_minimal()

```{r}
#barplot_layer

tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
  filter(!is.na(corefrag)) %>% 
  select(layer, corefrag) %>%
  ggplot() +
  aes(x = corefrag) +
  geom_bar() +  
  labs(x = "Core and Core fragments", y = "Frequency") +
  facet_wrap(~layer, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#Core-barplot_area

tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(!is.na(corefrag)) %>% # Exclude rows where corefrag is NA
  select(area, corefrag) %>%
  ggplot() +
  aes(x = corefrag) +
  geom_bar() +
  labs(x = "Core and Core fragments",
       y = "Frequency") +
  facet_wrap(~area, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

filtered_data <- tl_final %>%
  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
ggplot(data = filtered_data) + 
  aes(y = `tool-typology`) + 
  geom_bar() +
  labs(x = "Frequency",
       y = "Tool Type") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))

PCA-LAYER

```{r}
# prepare the data for PCA-layer

# Calculating Blank Frequency
blank_freq <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
  group_by(layer) %>%
  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')

# Calculating Core Frequency
core_freq <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
  group_by(layer) %>%
  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')

# Calculating Tool Diversity using Menhinick's Index
tool_div <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(tooltypegroup)) %>%
  group_by(layer) %>%
  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')

# Combining all the calculated frequencies into one tibble
final_table <- blank_freq %>%
  inner_join(core_freq, by = "layer") %>%
  inner_join(tool_div, by = "layer")

# Printing the final table
final_table

```{r}
# Blank Frequency
blank_freq <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
  group_by(layer) %>%
  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')

# Calculating Core Frequency
core_freq <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "e or d")) %>%
  group_by(layer) %>%
  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / n(), .groups = 'drop')

# Calculating Tool Diversity using Menhinick's Index
tool_div <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "e or d"), !is.na(tooltypegroup)) %>%
  group_by(layer) %>%
  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(sum(!is.na(tooltypegroup))), .groups = 'drop')

#------------------------------------------------------------------------------
# BM: I wasn't able to run any code below this line
##SS: fixed
# -----------------------------------------------------------------------------

# Calculating Retouch Frequency
retouch_freq <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
  group_by(layer) %>%
  summarise(retFreq = sum(retouch == 1, na.rm = TRUE) / n(), .groups = 'drop')

# Combining all the calculated frequencies, including retouch frequency, into one tibble
final_table <- blank_freq %>%
  inner_join(core_freq, by = "layer") %>%
  inner_join(tool_div, by = "layer") %>%
  inner_join(retouch_freq, by = "layer")

final_table

```{r}
# Joining calculated frequencies
final_table <- blank_freq %>%
  inner_join(core_freq, by = "layer") %>%
  inner_join(tool_div, by = "layer") %>%
  inner_join(retouch_freq, by = "layer")

# Converting layer to row names
final_table_rownames <- final_table %>%
  column_to_rownames(var = "layer")

```{r}
# Compute the PCA
library(FactoMineR)

res_pca <-
  PCA(final_table_rownames,
      graph = FALSE)

```{r}
library(factoextra)

# Visualize the contributions of variables to the first principal component
fviz_contrib(res_pca,
             axes = 1,   
             choice = "var")

```{r}
library(factoextra)

# contributions of variables to the second principal component
fviz_contrib(res_pca,
             axes = 2,   
             choice = "var")

```{r}
library(factoextra)

#  biplot
fviz_pca_biplot(res_pca)
```

```{r}
library(factoextra)
fviz_screeplot(res_pca)
```

```{r}
fviz_pca_var(res_pca)
```

# Run PCA
res_pca <- PCA(final_table_rownames, graph = FALSE)

# Visualize the contributions of variables to the first principal component
fviz_contrib(res_pca, axes = 1, choice = "var")

# Visualize the contributions of variables to the second principal component
fviz_contrib(res_pca, axes = 2, choice = "var")

# Visualize eigenvalues and explained variance
fviz_screeplot(res_pca)

# Create a biplot of the PCA
fviz_pca_biplot(res_pca)

# Visualize the variables on the principal components
fviz_pca_var(res_pca)

```{r}
# Run PCA
library(FactoMineR)
library(factoextra)

res_pca <- PCA(final_table_rownames, graph = FALSE)

# Get the coefficients of original variables for each principal component
loadings_matrix <- res_pca$var$coord

print(loadings_matrix)

```{r}
#Table

# Eigenvalues
eigenvalues <- res_pca$eig[1:2, 1]

# Variance percent
variance_percent <- res_pca$eig[1:2, 2]

# Cumulative variance percent
cumulative_variance_percent <- res_pca$eig[1:2, 3]

# Create the table
df <- data.frame(Eigenvalue = eigenvalues,
                 Variance_Percent = variance_percent,
                 Cumulative_Variance_Percent = cumulative_variance_percent)

rownames(df) <- c("Dim1", "Dim2")

print(df)

```{r}
#Table

library(gt)

# Create data frame
df <- data.frame(Eigenvalue = eigenvalues,
                 Variance_Percent = variance_percent,
                 Cumulative_Variance_Percent = cumulative_variance_percent)
rownames(df) <- c("Dim1", "Dim2")

# Create the table using gt
gt(df) %>%
  tab_header(
    title = "",
    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
  ) %>%
  cols_label(
    Eigenvalue = "Eigenvalue",
    Variance_Percent = "Variance (%)",
    Cumulative_Variance_Percent = "Cumulative Variance (%)"
  ) %>%
  fmt_number(
    columns = vars(Eigenvalue, Variance_Percent, Cumulative_Variance_Percent),
    decimals = 2
  )

feature frequency: d1 1 ta 1 tp1 0)

```{r}
library(dplyr)
library(FactoMineR)
library(tibble)

# Define estimated area and feature frequency data
area_data <- tibble(
  area = c("d1", "ta", "tp1"),
  feature_freq = c(1, 1, 0)
)

# Step 1: Number of artifacts by area
artifact_counts <- tl_final_area %>%
  group_by(area) %>%
  tally()

# Step 2: Calculate each variable

# Retouch Frequency 
retouch_freq <- tl_final_area %>%
  group_by(area) %>%
  summarise(retouchFreq = mean(!is.na(retouch)), .groups = 'drop')

# Blank Frequency
blank_freq <- tl_final_area %>%
  group_by(area) %>%
  summarise(blankFreq = mean(!is.na(blank)), .groups = 'drop')

# Core Frequency
core_freq <- tl_final_area %>%
  group_by(area) %>%
  summarise(coreFreq = mean(!is.na(`core-typology`)),.groups = 'drop')

# Lithic density (Artifact counts / estimated area)
lithic_density <- artifact_counts %>%
  left_join(area_data, by = "area") %>%
  mutate(lithic_density = n / estimated_area)

# Tool Diversity using Menhinick's Index
tool_div <- tl_final_area %>%
  group_by(area) %>%
  summarise(toolDiv = length(unique(tooltypegroup)) / sqrt(n()), .groups = 'drop')

# Step 3: Combining all variables

pca_final_table_area <- left_join(dimensions, retouch_freq, by = "area") %>%
  left_join(blank_freq, by = "area") %>%
  left_join(core_freq, by = "area") %>%
  left_join(lithic_density, by = "area") %>%
  left_join(tool_div, by = "area") %>%
  select(area, new_dimensions, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)

```{r}
pca_final_table_area <- pca_final_table_area %>%
    select(
        area, 
        estimated_area = estimated_area.x, 
        feature_freq = feature_freq.x, 
        retouchFreq, 
        blankFreq, 
        coreFreq, 
        n, 
        lithic_density, 
        toolDiv
    )

```{r}
# Step 4: Perform PCA
res_pca_area <- PCA(pca_final_table_area %>% select(-area, -n), graph = FALSE)

# Show the summary of PCA
print(res_pca_area)

print(res_pca_area$eig)
print(res_pca_area$var$coord)
```

```{r}
# TABLE: Eigenvalues and percentage of variance for each dimension of PCA 
library(kableExtra)

# Define the values manually (you would get these from your PCA results)
# BM: I prefer not to write values in the code like this. I think
# this is bad practice and leads to mistakes
eigenvalues <- c(4.411234, 2.588766)
prop_var_explained <- c(63.02, 36.98)
cumulative_var_explained <- c(63.02, 100.00)

# Create the data frame
results_df <- data.frame(
  Dimension = c("Dim. 1", "Dim. 2"),  # Only keep the dimensions you want to display
  Eigenvalue = eigenvalues,
  `Variance / percent` = prop_var_explained,
  `Cumulative / variance percent` = cumulative_var_explained
)

# Generate the table
kable(results_df,
      caption = "Table 3. Eigenvalues and percentage of variance for each dimension of PCA",
      align = 'c',
      col.names = c("Dimension", "Eigenvalue", "Variance (%)", "Cumulative Variance (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)

```{r}
#plotting Dim-1 and Dim-2
library(FactoMineR)
library(factoextra)

# Visualize the contributions of variables to the first principal component
fviz_contrib(res_pca_area, axes = 1, choice = "var")

# Visualize the contributions of variables to the second principal component
fviz_contrib(res_pca_area, axes = 2, choice = "var")

# having Dim1 and Dim2 plots next to each other

library(gridExtra)

# Create the first plot and assign it to p1
p1 <- fviz_contrib(res_pca_area, axes = 1, choice = "var")

# Create the second plot and assign it to p2
p2 <- fviz_contrib(res_pca_area, axes = 2, choice = "var")

# Use grid.arrange to place them next to each other
grid.arrange(p1, p2, ncol = 2)

loadings <- res_pca_area$var$coord

# 
eigenvalues <- res_pca_area$eig

#  Dim.1
contrib_dim1 <- (loadings[,1]^2 / eigenvalues[1, 1]) * 100

#  Dim.2
contrib_dim2 <- (loadings[,2]^2 / eigenvalues[2, 1]) * 100

contrib_df <- data.frame(
  Variable = rownames(loadings),
  Contrib_Dim1 = contrib_dim1,
  Contrib_Dim2 = contrib_dim2
)

print(contrib_df)

```{r}
# screeplot

fviz_eig(res_pca_area)

```{r}
# Correlation plots of variables for each of the tow relevant PCA dimensions

fviz_pca_var(res_pca_area,

             repel = TRUE) + 
  theme(text = element_text(size = 10))

```{r}
# Eigenvalues and percentage of variance
eig.val <- res_pca_area$eig
print(eig.val)

# Variable coordinates (correlations with principal components)
var.coord <- res_pca_area$var$coord
print(var.coord)

```{r}
#PCA biplot for Dimensions 1 and 2

fviz_pca_biplot(res_pca_area, labelsize = 3, repel = TRUE)

```{r}
# biplot 
p <- fviz_pca_biplot(res_pca_area, label = "none", 
                     repel = TRUE, labelsize = 4)

p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
                  y = res_pca_area$ind$coord[, 2], 
                  label = pca_final_table_area$area), 
             size = 3, nudge_x = 0.1, nudge_y = 0.1)

```{r}
individual_coordinates = res_pca_area$ind$coord
variable_coordinates = res_pca_area$var$coord

```{r}
# For variables
fviz_cos2(res_pca_area, choice = "var")

# For individuals
fviz_cos2(res_pca_area, choice = "ind")

DONE WITH PCA

individual_coordinates <- res_pca_area$ind$coord

individual_coordinates_df <- as.data.frame(individual_coordinates)
individual_coordinates_df$area <- unique(area_data$area)

print(individual_coordinates_df)

# Extract individual coordinates
individual_coordinates <- res_pca_area$ind$coord

# Convert to a data frame for better visibility and add the 'area' column back for context
individual_coordinates_df <- as.data.frame(individual_coordinates)
individual_coordinates_df$Area <- unique(area_data$area)  # Make sure the order matches

# Use kable to create a beautiful table
kable(individual_coordinates_df, 
      col.names = c("Area", "Principal Component 1", "Principal Component 2"), 
      caption = "PCA Coordinates for the Archaeological Areas excavation",
      align = c('l', 'c', 'c'),
      digits = 3) %>%
  kable_styling("striped", full_width = F)

```{r}
# Get the loadings (coefficients) of original variables for each principal component
loadings_matrix <- res_pca$var$coord

```{r}
# Eigenvalues
eigenvalues <- res_pca$eig[1:2, 1]
variance_percent <- res_pca$eig[1:2, 2]
cumulative_variance_percent <- res_pca$eig[1:2, 3]

# Create the table
df <- data.frame(Eigenvalue = eigenvalues,
                 Variance_Percent = variance_percent,
                 Cumulative_Variance_Percent = cumulative_variance_percent)

rownames(df) <- c("Dim1", "Dim2")

# Print the table
print(df)

```{r}
library(gt)

gt(df) %>%
  tab_header(
    title = "",
    subtitle = "Eigenvalues and percentage of variance for each dimension of PCA"
  ) %>%
  cols_label(
    Eigenvalue = "Eigenvalue",
    Variance_Percent = "Variance (%)",
    Cumulative_Variance_Percent = "Cumulative Variance (%)"
  ) %>%
  fmt_number(
    columns = c("Eigenvalue", "Variance_Percent", "Cumulative_Variance_Percent"),
    decimals = 2
  )

>>>>

```{r}
# Generate a summary table by grouping by 'depth' and 'area'
summary_table_depth_tool_type <- tl_final %>%
  group_by(area, depth) %>%
  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE), .groups = 'drop')

# Create a full summary table by joining it back to the original summary table based on 'depth' and 'area'
summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
  left_join(summary_table_depth_tool_type, by = c("area", "depth"))

# Remove the existing 'tool_type' if it exists
if ("tool_type" %in% names(summary_table_depth_area_layer)) {
  summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
    select(-tool_type)
}

# Join again based on 'depth', this time being specific about what we want
summary_table_depth_area_layer <- 
  summary_table_depth_area_layer %>%
  left_join(summary_table_depth_tool_type %>% select(depth, tool_type), 
            by = "depth")

@@

google_sheet_url <- 
  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
sum_depth_sheet <- read_sheet(google_sheet_url)

@@

```{r}
# Create the plot: WABI_Level_logscale
# here I do not consider those levels with a value 0 of retouch frequency. 
filtered_data <- sum_depth_sheet %>% 
  filter(lithic_dens > 0 & retouch_freq > 0)

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  geom_text(aes(label = label), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 4) +
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

@@

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, I consider the thickness; is that OK?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
##core frequency
##blank frequency
##retouch frequency
##tool diversity
##thickness of deposit (???)
##lithic density

@@

pca_data <- sum_depth_sheet %>% 
  select(lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
  na.omit()  # Remove rows with NA values

@@

```{r}
# performin PCA

pca_data <- sum_depth_sheet[, c('lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]

@@

```{r}
eigen_df <- data.frame(
  eigenvalue = res.pca$eig[,1],
  `percentage of variance` = res.pca$eig[,2],
  `cumulative percentage of variance` = res.pca$eig[,3]
)

@@

# Re-order the columns based on the corrected names
eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage.of.variance", "cumulative.percentage.of.variance")]

@@

```{r}
# Visualize biplot:1
fviz_pca_biplot(res.pca,
                axes = c(1, 2),
                labelsize = 2
                )
rownames(res.pca$ind$coord) <- sum_depth_sheet$label

@@

```{r}
# Visualize biplot: 2
fviz_pca_biplot(res.pca,
                axes = c(3, 4),
                labelsize = 2
                )
rownames(res.pca$ind$coord) <- sum_depth_sheet$label

@@
